---
title: "Species_distribution_Hporpoise"
author: "Jo-Hannes Nowé (VLIZ)"
format: 
  html: 
    theme: yeti
editor: visual
toc: true
number-sections: true
highlight-style: pygments
bibliography: references.bib
csl: apa.csl
---

```{r folder_structure}
#| echo: false
# Define different folders
downloaddir <- "data/raw_data"
datadir     <- "data/derived_data"
envdir <-"data/raw_data/environmental_layers"
occdir <-"data/raw_data/occurrences"
spatdir <- "data/raw_data/spatial_layers"
folderstruc <- c(downloaddir,
                 datadir,
                 envdir,
                 occdir,
                 spatdir)

#Check for their existence, create if missing
for(i in 1:length(folderstruc)){
  if(!dir.exists(folderstruc[i])){
    # If not, create the folder
    dir.create(folderstruc[i],recursive = TRUE)
    cat("Folder created:", folderstruc[i], "\n")
  } else {
    cat("Folder already exists:", folderstruc[i], "\n")
  }
}
```

```{r load_packages}
#| results: false
#| warning: false
#| error: false
#Load the necessary R packages
package_list <- c("arules",
                  "arrow",
                  "arulesViz",
                  "BiocManager",
                  "CoordinateCleaner",
                  "dismo",
                  "doParallel",
                  "downloader",
                  "earth",
                  "foreach",
                  "ks",
                  "mgcv",
                  "ows4R",
                  "ranger",
                  "raster",
                  "Rarr",
                  "sdm",
                  "sf",
                  "sp",
                  "spatialEco",
                  "stacks",
                  "stats",
                  "stars",
                  "terra",
                  "tidymodels",
                  "tidyverse",
                  "utils",
                  "worrms",
                  "xgboost")
#For the packages that need to be installed from github
package_list_github <-c("vlizBE/imis")
invisible(lapply(package_list, library, character.only = TRUE,quietly=TRUE))
library(imis)
```

# Introduction

In this notebook we fit a species distribution model to EurOBIS data in the EDITO environment. By altering the choices in the first code chunk, this notebook can be run for a species, study area or time period of choice. This workflow is being updated as improvements are made.

There are 3 main parts in the workflow: the collection of data from the data lake, model training and making predictions.

# Collection of data out of the data lake

In the first part of the analysis, we are setting up the data frame on which we want to build the model. In short we want to have occurrences and background points coupled with their respective values for temperature, salinity, net primary production and bathymetry: the predictors used in the model.

## Defining user choices

First, some choices are made on the species, spatial and temporal extent of the study. By altering this part of the code, you can train a model for your use case.

```{r}
region <- c('II','III') #ospar region to define the study area
aphia_id <-137117   #this is Phocoena phocoena
word_filter <- c("stranding", "museum") #to filter out datasets based on a keyword
date_start <- "1999-01-01T00:00:00"
date_end <- "2019-01-01T00:00:00"
temporal_extent <- lubridate::interval(date_start,date_end)
```

## Presence data

### Defining the study area

```{r}
#Download and unzip the ospar shapefiles
url <- "https://odims.ospar.org/public/submissions/ospar_regions/regions/2017-01/002/ospar_regions_2017_01_002-gis.zip"
download.file(url,file.path(spatdir,"ospar_REGIONS.zip"),mode="wb")
unzip(zipfile=file.path(spatdir,"ospar_REGIONS.zip"),exdir=spatdir)

#Visualize the different regions
ospar<- st_read(file.path(spatdir,"ospar_regions_2017_01_002.shp"))

ospar <- ospar[ospar$Region %in% region,]
ggplot(data=ospar)+geom_sf()
#Only keeping the region and geometry info
ospar<- ospar %>% dplyr::select(Region)

#Bring together the different ospar regions into one area
spatial_extent <- st_union(ospar)

```

### Downloading occurrence data from EurOBIS

The occurrence data we use comes from EurOBIS. Using the arrow R package, we can pre-filter and manipulate the data before collecting, making these steps fast and efficient.

```{r eurobis_connection}
acf <- S3FileSystem$create(
  anonymous = T,
  scheme = "https",
  endpoint_override = "s3.waw3-1.cloudferro.com"
)
eurobis <- arrow::open_dataset(acf$path("emodnet/biology/eurobis_occurence_data/eurobisgeoparquet/eurobis_no_partition_sorted.parquet"))
```

```{r eurobis_download}
bbox <- sf::st_bbox(spatial_extent)
mydata.eurobis <- eurobis %>%
  filter(aphiaidaccepted==aphia_id,
         longitude > bbox[1], longitude < bbox[3],
         latitude > bbox[2], latitude < bbox[4],
         observationdate >= as.POSIXct(date_start),
         observationdate <= as.POSIXct(date_end)) %>%
  dplyr::select(datasetid,
                latitude,
                longitude,
                time=observationdate,
                scientific_name = scientificname_accepted,
                occurrence_id=occurrenceid) %>%
  mutate(year=year(time),
         month=month(time),
         day = day(time))%>%
  collect()%>%
  sf::st_as_sf(coords=c("longitude", "latitude"),
               crs=4326)

```

### Filtering and processing the data

With the custom function `fdr2`from @hermanpmj2020, we can generate metadata on the list of datasets we use the occurrences from. We use this information to further filter the data.

```{r metadata_function}
#| include: false
#function to read dataset characteristics, code from: https://github.com/EMODnet/EMODnet-Biology-Benthos_greater_North_Sea

fdr2<-function(dasid){
  datasetrecords <- datasets(dasid)
  dascitations <- getdascitations(datasetrecords)
  if(nrow(dascitations)==0)dascitations<-tibble(dasid=as.character(dasid),title="",citation="")
  if(nrow(dascitations)==1) if(is.na(dascitations$citation)) dascitations$citation<-""
  daskeywords <- getdaskeywords(datasetrecords)
  if(nrow(daskeywords)==0)daskeywords<-tibble(dasid=as.character(dasid),title="",keyword="")
  if(nrow(daskeywords)==1) if(is.na(daskeywords$keyword))daskeywords$keyword<-""
  dascontacts <- getdascontacts(datasetrecords)
  if(nrow(dascontacts)==0)dascontacts<-tibble(dasid=as.character(dasid),title="",contact="")
  if(nrow(dascontacts)==1) if(is.na(dascontacts$contact))dascontacts$contact<-""
  dastheme <- getdasthemes(datasetrecords)
  if(nrow(dastheme)==0)dastheme<-tibble(dasid=as.character(dasid),title="",theme="")
  if(nrow(dastheme)==1) if(is.na(dastheme$theme))dastheme$theme<-""
  dastheme2 <- aggregate(theme ~ dasid, data = dastheme, paste, 
                         collapse = " , ")
  daskeywords2 <- aggregate(keyword ~ dasid, data = daskeywords, 
                            paste, collapse = " , ")
  dascontacts2 <- aggregate(contact ~ dasid, data = dascontacts, 
                            paste, collapse = " , ")
  output <- dascitations %>% left_join(dascontacts2, by = "dasid") %>% 
    left_join(dastheme2, by = "dasid") %>% left_join(daskeywords2, 
                                                     by = "dasid")
  return(output)
}
```

```{r metadata_creation}
datasetidsoi <- mydata.eurobis %>% 
  sf::st_drop_geometry() %>%
  distinct(datasetid) %>% 
  mutate(datasetid = as.numeric(str_extract(datasetid, "\\d+")))
#==== retrieve data by dataset ==============
all_info <- data.frame()
for (i in datasetidsoi$datasetid){
  dataset_info <- fdr2(i)
  all_info <- rbind(all_info, dataset_info)
}
names(all_info)[1]<-"datasetid"
write.csv(all_info,file=file.path(datadir,"allDatasets.csv"),row.names = F, append=FALSE)
alldataset <- read.csv(file.path(datadir,"allDatasets.csv"))
```

We filter the datasets based on keywords such as "museum" and "strandings".

```{r filter_dataset}
alldataset <- alldataset %>%
  rowwise() %>%
  mutate("discard"=any(across(-description, ~grepl(paste(word_filter, collapse = "|"), .,ignore.case=TRUE))))

alldataset_selection<-alldataset%>%
  filter(discard==FALSE)

alldataset_flagged <- alldataset%>%
  filter(discard==TRUE)
```

We also filter out duplicate samples based on coordinates and time of measurement.

```{r filter_occurrences}
mydata.eurobis <- mydata.eurobis %>%
  filter(!is.na(time))%>%
  st_filter(y = spatial_extent)%>%
  filter(datasetid %in% alldataset_selection$datasetid)%>%
  dplyr::distinct(occurrence_id,.keep_all = TRUE)%>%
  arrange(time)%>%
    mutate(year_month=paste(year,month,sep='-'))%>%
  mutate(year_month=factor(year_month,levels=unique(year_month),ordered=TRUE))%>%
  dplyr::mutate(longitude = sf::st_coordinates(.)[,1],
                latitude = sf::st_coordinates(.)[,2],
                occurrence_status = 1)%>%
  dplyr::select(!c(datasetid,occurrence_id))%>%
  sf::st_drop_geometry()

#Remove duplicates
mydata.eurobis <- cc_dupl(mydata.eurobis, lon = "longitude", lat = "latitude",value = "clean",species="scientific_name", additions="time")
```

## Background data

The background data is created using the target-group background method, where the occurrences of other species in the same target group are used to infer the sampling effort [@phillips2009sample].

```{r target_group_choices}
species <- worrms::wm_id2name(aphia_id)
classification <- wm_classification(aphia_id) 
class <- which(classification[,2]=="Class")
target_group <- classification[[class,3]]
```

Occurrences from the target group in the same datasets as our occurrences are downloaded.

```{r download_target_group}
list_dasid <- alldataset_selection$datasetid

#Some workaround to get the Class information while it is not in the parquet file
#Generate a list of all the distinct aphiaids over the datasets
aphiaid_list <- eurobis %>%
  filter(datasetid %in% list_dasid,
         longitude > bbox[1], longitude < bbox[3],
         latitude > bbox[2], latitude < bbox[4],
         observationdate >= as.POSIXct(date_start),
         observationdate <= as.POSIXct(date_end)) %>%
  dplyr::select(aphiaid) %>%
  distinct()%>%
  collect()

#Check which of the aphiaids belong to the target group
target_aphiaids <- worrms::wm_classification_(aphiaid_list[[1]])%>%
  dplyr::filter(rank == "Class",
                scientificname==target_group)%>%
  dplyr::select(aphiaid="id")%>%
  dplyr::mutate(aphiaid=as.numeric(aphiaid))

target_background <- eurobis %>%
  filter(datasetid %in% list_dasid,
         aphiaidaccepted %in% target_aphiaids$aphiaid,
         longitude > bbox[1], longitude < bbox[3],
         latitude > bbox[2], latitude < bbox[4],
         observationdate >= as.POSIXct(date_start),
         observationdate <= as.POSIXct(date_end)) %>%
  dplyr::select(datasetid,
                latitude,
                longitude,
                time=observationdate,
                scientific_name = scientificname_accepted,
                occurrence_id = occurrenceid) %>%
  mutate(year=year(time),
         month=month(time),
         day = day(time))%>%
  collect()%>%
  sf::st_as_sf(coords=c("longitude", "latitude"),
               crs=4326)

target_background <- target_background %>%
  filter(!is.na(time))%>%
  st_filter(y = spatial_extent)%>%
  dplyr::distinct(occurrence_id,.keep_all = TRUE)%>%
  arrange(time)%>%
    mutate(year_month=paste(year,month,sep='-'))%>%
  mutate(year_month=factor(year_month,levels=unique(year_month),ordered=TRUE))%>%
  dplyr::mutate(longitude = sf::st_coordinates(.)[,1],
                latitude = sf::st_coordinates(.)[,2],
                occurrence_status = 0)%>%
  dplyr::select(!c(datasetid,occurrence_id))%>%
  sf::st_drop_geometry()
```

```{r}
pa_occurrence <- rbind(mydata.eurobis, target_background)
summary(pa_occurrence)
```

## Environmental data layers

We collect the environmental data corresponding to the presence/background points, which is stored in .zarr format in the datalake. This is done by using the editoTools.

```{r}
source("https://raw.githubusercontent.com/EDITO-Infra/R_tools/main/editoTools.R")
```

We define our environmental variables: temperature, salinity, net primary production and bathymetry.

```{r}
parameters <- list(
  "thetao"= c("par" = "thetao", "fun" = "mean"),
  "so"= c("par" = "so", "fun" = "mean"),
  "npp"= c("par" = "npp", "fun" = "mean", "convert_from_timestep" = "86400000"),
  "elevation"=c("par"= "elevation", "fun"="mean"))
```

```{r}
#Takes +- 1 min for thetao and so, 45min for npp, 5 for bathymetry
df_occ <- enhanceDF(inputPoints = pa_occurrence %>% 
                            mutate(Time = as.POSIXct(paste(year,month,1,sep = "-")))%>%
                                     dplyr::select(Time,Longitude=longitude,Latitude=latitude,occurrenceStatus=occurrence_status),
                          requestedParameters = parameters,
                          requestedTimeSteps = NA,
                          stacCatalogue = EDITOSTAC,
                          verbose="",
                          select_layers=rep(1,4))
df_occ <- df_occ %>% na.omit()
```

# Model training

We use the data frame with environmental values as input data for the model. The model is trained using the tidymodels R package.

```{r}
#Turn our occurrence into a factor so that there are two known levels, either presence or absence.
df_occ$occurrenceStatus<-factor(df_occ$occurrenceStatus)

data <- df_occ%>%
  dplyr::select(occurrenceStatus,thetao,so,npp,elevation)%>%
  drop_na()%>%
  as_tibble()
```

```{r}
# Data splitting
set.seed(222)
#Put 4/5 in the training set
data_split<- initial_split(data, prop=4/5)

#Create data frames for the two sets:
train_data <- training(data_split)
test_data <- testing(data_split)

#Create folds in the training data so we can do our validation.
#Out of the 80% reserved for training data, use 10% each time for validation
#This is why we choose the number of folds v=8
folds <- vfold_cv(train_data, v=8)
folds

Occurrence_rec <- 
  recipe(occurrenceStatus ~., data=train_data)%>%
  step_normalize(all_numeric_predictors())
```

```{r}
#Because we use tune_grid()
ctrl_grid <- control_stack_grid()
ctrl_res <- control_stack_resamples()


#Unregister the parallel nodes, parallelised in the editoTools
unregister <- function() {
  env <- foreach:::.foreachGlobals
  rm(list=ls(name=env), pos=env)
}
unregister()
```

First a Random Forest, Generalized Additive Model (GAM), MARS and XGBoost model are fitted on the training data. Afterwards the most optimal combination of the model configurations is put together into an ensemble model.

```{r}
rf_mod <- rand_forest(trees=500,mode="classification",engine="ranger")

rf_wf <-
  workflow() %>%
  add_model(rf_mod) %>%
  add_recipe(Occurrence_rec)

rf_fit <-
  fit_resamples(
    rf_wf,
    resamples = folds,
    control = ctrl_res
  )
```

```{r}
gam_mod <- gen_additive_mod(adjust_deg_free = tune()) %>% 
  set_engine("mgcv") %>% 
  set_mode("classification")
gam_wf <-
  workflow()%>%
  add_recipe(Occurrence_rec)%>%
  add_model(gam_mod, formula = occurrenceStatus ~ s(elevation)+ s(thetao)+ s(so) + s(npp))
gam_grid <- grid_regular(adjust_deg_free(),
                         levels=4)
gam_fit <-
  gam_wf%>%
  tune_grid(
    resamples = folds,
    grid=gam_grid,
    control=ctrl_grid
  )
```

```{r}
xgb_mod <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")
xgb_wf <- workflow() %>%
  add_model(xgb_mod) %>%
  add_recipe(Occurrence_rec)

xgb_grid <- parameters(trees(),
                       tree_depth(c(3,20)),
                       learn_rate(c(-4,0)))
xgb_grid <- grid_regular(xgb_grid,
                         levels=c(trees=5,tree_depth=3,learn_rate=3))
xgb_fit <- 
  xgb_wf %>% 
  tune_grid(
    resamples = folds,
    grid = xgb_grid,
    control=ctrl_grid)
```

```{r}
mars_mod <- 
  mars(prod_degree = tune(), prune_method = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("earth")#,glm=list(family=binomial))
mars_mod
mars_grid <- grid_regular(prod_degree(),prune_method(),
                          levels=c(prod_degree=2,prune_method=2))
mars_wf <-
  workflow()%>%
  add_model(mars_mod)%>%
  add_recipe(Occurrence_rec) 
mars_fit <-
  mars_wf%>%
  tune_grid(
    resamples = folds,
    grid=mars_grid,
    control=ctrl_grid
  )
```

```{r}
#Stacking the individual models into an ensemble model

stack_data <- 
  stacks() %>%
  add_candidates(rf_fit) %>%
  add_candidates(mars_fit)%>%
  add_candidates(gam_fit)%>%
  add_candidates(xgb_fit)

stack_mod <-
  stack_data %>%
  blend_predictions()

#Fitting the model on the whole training set
stack_fit <-
  stack_mod %>%
  fit_members()
stack_fit #the final model
```

# Making monthly predictions

Using the `getRasterSlice` from the `editoTools` we collect monthly raster layers to make predictions on habitat suitability for the study area in that month.

```{r}
monthly_dates <- seq(from= as.POSIXct(date_start), to= as.POSIXct(date_end), by = "month")
variables <- rep(names(parameters),each=length(monthly_dates))
layer_names <- tibble("month"=rep(monthly_dates,length(parameters)))%>%
  mutate(variable= variables)
raster_list<-map2(layer_names$variable,layer_names$month, \(variable,month) getRasterSlice(variable,
                                      stacCatalogue = EDITOSTAC,
                                                  lon_min = bbox[1],
                                      lon_max = bbox[3],
                                      lat_min = bbox[2],
                                      lat_max = bbox[4],
                                      requestedTimeSteps = NA,
                                      date = month,
                                      select_layers = 1))

thetao <- raster_list[1:length(monthly_dates)]
so <- raster_list[(length(monthly_dates)+1):(2*length(monthly_dates))]
npp <- raster_list[(2*length(monthly_dates)+1):length(raster_list)]


bathy_url <- EDITOSTAC[EDITOSTAC$par=="elevation","href"][4,]
new_bathy_url <- sprintf('ZARR:"/vsicurl?list_dir=no&retry_delay=60&max_retry=3&url=%s":/%s',bathy_url,"elevation")
bathy_raster <- terra::rast(new_bathy_url)
```

```{r}
prediction <- predict(stack_fit,test_data)%>%
  bind_cols(predict(stack_fit,test_data, type="prob")) %>%
  bind_cols(test_data %>%
              dplyr::select(occurrenceStatus))
prediction
accuracy_vec(prediction$occurrenceStatus,prediction$.pred_class)
```

```{r}
#Make a custom function that can be used with the terra::predict function
predprob <- function(...) predict(...,type="prob")$.pred_1
```

```{r}

#Load the different layers (see environmental clean script)
thetao <- terra::rast(thetao)
so <- terra::rast(so)
npp <- terra::rast(npp)
#Different layers, crop them to the same extent of bathymetry raster
thetao <- crop(thetao,bathy_raster) 
so <- crop(so,bathy_raster)
npp <- crop(npp,bathy_raster)
ext(bathy_raster) <- ext(thetao)
bathy_raster <- resample(bathy_raster, thetao)
monthly_prediction <- list()
#Loop over every monthly layer
for(i in 1:nlyr(thetao)){
  monthly_info <- c(thetao[[i]],so[[i]],npp[[i]],bathy_raster)
  #Delete the NA values
  mask <- !is.na(monthly_info)
  monthly_info <- terra::mask(monthly_info,mask)
  #To make sure the names of the layers correspond to the names the model was trained on
  names(monthly_info) <- c(
    "thetao",
    "so","npp","elevation")
  monthly_prediction[[i]] <- predict(monthly_info,stack_fit,fun=predprob,na.rm=TRUE)
}
#To turn it into a raster layer  
monthly_prediction <- rast(monthly_prediction)
```

```{r}
writeCDF(monthly_prediction,file.path(datadir,paste0("monthly_predictions_",aphia_id,".nc")))
```

# References
